---
title: 'Social Media Analytics VI'
subtitle: "Synergie of practical 1-5"
output:
  html_document:
    theme: united
    highlight: tango
    css: custom.css
    toc: true
    toc_depth: 2
editor_options:
  chunk_output_type: console
---

```{r opts, echo = FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
```

VU Amsterdam

# This practical

In the past 5 practicals you have learned a range of skills in the area of content analysis.
However, due to the extended, bi-weekly format of this first part of the course, it might not immediately be obvious to you how all of these skills fit together.
In this practical we'll look back on what you learned, and show how it all comes together as a skillset for **building and validating an automatic content analysis tool**.


## What have you learned?

Here is a quick summary of key takeaways from practical 1 to 5:

1. **Manual Content Analysis** is a technique for making **valid** and **reliable** inferences from texts
2. How to achieve and measure **intercoder reliability** (Cohen's Kappa, Krippendorfs Alpha)
3. Performing **Automatic Content Analysis** (using search queries), and how to measure the **validity** (precision and recall)
4. Working with textual data in R and RStudio
5. How to perform Automatic Content Analysis using Large Language Models using **Zero-shot classification**

The goal of these practicals was to teach you how to perform **automatic content analysis** at a professional level. **Zero-shot classification** is a powerful technique that empowers you to analyze social media content at scale. But in order to use this technique properly, you also need to know how to create a valid and reliable **gold standard** using **manual content analysis**. 

In this practical we'll be putting all of this together. We'll start with a manual content analysis dataset and assess it's intercoder-reliability. Then we'll create a function for zero-shot classification, and assess its precision and recall based on the manual content analysis data. Finally, we'll attempt to improve the precision and recall by trying different prompts.      


## Why is this important for Social Media Analytics?

Social media provide a goldmine of information about society, but **manual content analysis is too slow and expensive** to use this information effectively. Companies want to be able to continuously monitor how people talk about them. Researchers want big data for comparative and longitudinal research. **Automatic content analysis** enables us to perform content analysis at a big scale, but it used to be quite difficult to automate many types of content-analysis with acceptable validity. The state-of-the-art **zero-shot classification** technique that you learn in this course offers a flexibility and performance that researchers and professionals several years ago could only dream of. 

Consider the homework literature of last week, in which Atteveldt et al., analyzed the validity of **sentiment analysis** techniques using dictionaries and machine learning approaches. **Dictionaries** were relatively easy to use, but they performed very poorly in terms of validity. The best results were achieved using complex **Supervised Machine Learning** approaches that required you to first train (i.e. supervise) a model using a lot of examples. But this meant that you still needed to perform a lot of manual content analysis in order to train your model, and these models could only be used for the specific task that you trained them on. Furthermore, many of these traditional techniques only worked with textual data, and only for specific languages.  

The **zero-shot** classification technique provides a powerful alternative. You can ask any question without having to first train the model. In this course we only use this for analyzing textual content in English, but the same technique also works for multimedia and multilingual content. This is especially important for social media data, which today is highly diverse and multimodal.  


## Today's assignment

In last week's homework literature, you read the article by [Van Atteveldt et al., (2021)](https://www.tandfonline.com/doi/pdf/10.1080/19312458.2020.1869198)--presumably with great interest and maybe even twice!
In this article, they analyze the validity of sentiment analysis techniques, comparing manual annotation, crowd-coding, dictionary approaches and machine-learning approaches. 

They draw the following conclusions (abstract):

```
The three main conclusions of this article are that: 
(1) The best performance is still attained with trained human or crowd coding; 
(2) None of the used dictionaries come close to acceptable levels of validity; and 
(3) machine learning, especially deep learning, substantially outperforms dictionary-based methods but falls short of human performance. 

From these findings, we stress the importance of always validating automatic text analysis methods before usage.
```

This conclusion implies that for automatic content analysis we would need to use machine learning approaches, and meticulously train and validate them using manual content analysis. 

What is fascinating about this paper, is that even though it is fairly recent (2021), it does not mention Large Language Models (LLM) at all. THATS HOW FAST THE FIELD HAS EVOLVED. Zero-shot classification wasn't a thing yet. 

So what we'll do in this tutorial is to see how well we can approach or even exceed the performance of the best machine learning models in the study using zero-shot classification. Even though the analyses in Van Atteveldt et al., were performed by experts with much more time on their hands, You have technological advancement on your side! **Can you, having only recently learned about content analysis at all, beat the recent state-of-the-art using this new method?**
  


# 1. Creating a Gold Standard

Van Atteveldt et al., had to create a gold standard to evaluate and compare the different content analysis approaches. For this, they conducted a manual content analysis of 300 units that were coded by 3 expert coders (the authors themselves).
If they can reliably code the sentiment of economic headlines the same way, it would indicate that their measurement is reliable (and valid).

The authors used Krippendorf's alpha as the reliability metric.
In **practical 2**, you manually calculed Cohen's Kappa because Krippendorf's alpha is difficult to compute by hand. 
But this time we're working in R, where we can use the `irr` package to compute Krippendorf's alpha for us.

## Import the Content Analysis data into R

We'll first need to read the manually coded data into R.
When you'll be doing your own manual content analysis in the project in period 6, we'll recommend you to work in Google Sheets. So to practice importing this type of data, we prepared the manual content analysis data from the study in [this google sheet](https://docs.google.com/spreadsheets/d/18oquhpDp3uyRzP0jcLbkBXOUKoLcCtRo8m73AodKnFM/edit?usp=sharing). 

When you open the Google Sheet link, you'll see that every row represents a headline, and for every headline there are three columns with sentiment coding by different coders (`coder_1_sentiment`, etc).
Notice that there are 4 sheets (the tabs at the bottom of the page).
Each coder has independently coded the sentiment in a separate sheet, to avoid being influenced by each others' answers.
The results where then put together in the **combined** sheet. 
Because the independently coded the sentiment, we can now use this data to compute the intercoder reliability.

To import the Google Sheet data into R, we'll first export the data from Google Sheets to a CSV file.^[There is also an R package for directly reading data from Google Sheets, but doing it manually is a more useful exercise]. Simply go to `File -> Download -> Comma separated values (.csv)`. Note that you can only export the current sheet to CSV, so **Make sure (!!)** to select the **combined** sheet to download the combined data.

**If you're working on JupyterHub**, you first need to upload the file. You can easily do this with the upload button in the Files tab (bottom-right corner of RStudio).

Now that you downloaded the file to your computer (probably your downloads/ folder) you can read it into R. 


```{r}
library(tidyverse)

## Make sure to change the file location if it's different for you
intercoder <- read_csv("~/Downloads/intercoder_reliability - combined.csv")
```

## Calculate the Krippendorf's Alpha

Now we can calculate the Krippendorf's Alpha using the `irr` package (which stands for InterRater Reliability, but it's the same thing).
The `kripp.alpha` function requires us to give it the data as a matrix where the rows are coders and the columns are units. 
So in order to create this data, we'll 

* first **select** only the coder columns (from `coder_1_sentiment` up to column `coder_3_sentiment`). 
* then **transpose** (`t()`) the data, which swaps the rows and columns (so that the rows become the coders and the columns the units).


```{r, results='hide'}
library(irr)

intercoder |>
  select(coder_1_sentiment:coder_3_sentiment) |>
  t() |>
  kripp.alpha(method = 'interval')
```

## Assignment 1

> **1.1**. Perform the steps above. In your answer, show the code you used and the output of the kripp.alpha function.

> **1.2**. Interpret what the Krippendorf's Alpha tells us about the relibility of this manual content analysis data. Does it match what is reported in the article?

> **1.3**. We called the kripp.alpha function with `method = 'interval'`. Explain what this does and what the alternatives are. Is this the correct choice for this data?

> **1.4**. The `combined` data also has a single `gold` (gold standard) column that represents what was decided to be the correct sentiment. How did Van Atteveldt et al., decide this, and why are there missing values (NAs)?


# 2. Building and validating a zero-shot classifier

Let's first clean up the manual content analysis data so that we just have the headlines with the gold value.
We'll call this data `gold`, because it's our gold standard moving forward.

```{r}
gold <- intercoder |>
  select(id, headline, gold) |>
  filter(!is.na(gold))

gold
```

The process of building and validating our zero-shot classifier is now quite straighforward:

1. We'll create our own sentiment classifier function just like in practical 5.
2. Then we'll use this function to classify the sentiment for all the headlines in the gold standard.
3. We'll use the gold standard to calculate the precision and recall of our classifier


## Creating a sentiment classifier function

Last time we used Gemini and Ollama.
In order to get the best performance, we'll use Gemini again this time.
But when classifying around 300 headlines you would run into the limits of your free Gemini API key.
For this assignment, you should therefore use a paid API key.
No worries though, because you can use ours!
You can find it on Canvas under this week's module.

### Setting up Gemini in TidyLLM

The process is identical to before. Simply set the key once, and then we can use `tidyllm` with gemini_chat. Make sure to use the `gemini-2.5-flash-lite` model for this assignment to avoid running into rate limits (aka making your own life harder). 

```{r, eval=F}
library(tidyllm)
Sys.setenv(GOOGLE_API_KEY = "[ENTER API KEY HERE]")

llm_message("Hi, how are you?") |>
  gemini_chat(.model = "gemini-2.5-flash-lite")%>%
  get_reply()
```

### Creating a classifier function

Now let's build our classifier function, similar to how we did this in practical 5.
This time, we'll give our model 2 parameters: the text to classify and the prompt for instructing the model how to code sentiment. 
Our default instruction is just a plain instruction without a clear definition of sentiment.
By making instruction a parameter, we can later change the instruction to more clearly instruct the model what we mean by sentiment.


```{r}
sentiment_classifier <- function(text, instruction = "Code the sentiment of this text") {
  sentiment_schema = tidyllm_schema(
    sentiment = field_fct("The sentiment of the text", c("positive", "neutral", "negative"))
  )
  
  task <- "You are a coder in a content analysis study. Your task is to code the sentiment of the given text, strictly based on the following instructions of the researcher."
  
  result <- llm_message(task) |>
    llm_message(paste("This is the instruction:", instruction)) |>
    llm_message(paste("This is the text:", text)) |>
    gemini_chat(.model = "gemini-2.5-flash-lite", 
                .json_schema = sentiment_schema,
                .temperature = 0) |>
    get_reply_data() 
  
  return(result$sentiment)
}
```

Ok, let's see if it works!

```{r, eval=F}
sentiment_classifier("This is a great example")
sentiment_classifier("This is a terrible example")
```

Let's see if we can change the outcome by changing the instruction.
Perhaps you recall that in the first lecture we showed you multiple news headlines, and everyone had to say whether they believe the headline was positive, negative or neutral.
Here you clearly saw that many headlines allowed for different interpretations, depending on how you define sentiment.
The lessen here was that if we do not properly define sentiment, different coders will have different interpretations, leading to low `intercoder reliability`.
To make sure every coder interprets sentiment in roughly the same way, we provide a codebook that guides their interpretations.
For a Large Language Model, we can similarly provide instructions for how to interpret sentiment.

One example that we showed was a variation on the classic philosophical [trolley problem](https://en.wikipedia.org/wiki/Trolley_problem). 
Here we have another variation. 
Let's see if we can change the interpretation of the model with a simple instruction.

```{r, eval=F}
headline = "A surgeon killed one involuntary organ donor to save five people"

sentiment_classifier(headline)

sentiment_classifier(headline, instruction = "Your task is to code sentiment from a utilitarian perspective that maximizes utility")
```

When I ran this code, adding this instruction indeed changed the outcome from negative to positive (there is a small chance you get different results, but rest assured that the process works). By telling the model to code sentiment from a utilitarian perspective, we tell it to interpret sentiment based on the net positive outcome, regardless of the morally questionable means to this end.

Of course, this is a rather silly example, and a single example like this should not be taken as proof that our instruction generalizes to other headlines. 
It is possible that the main reason why our instruction worked for this example is that the model simply recognized the trolley problem, which is strongly related to utilitarianism.
To validate whether a instruction is good, you need to test is for a large number of examples (which we'll do in assignment 3).

## Assignment 2

Here is another example of a news headline with multiple interpretations of sentiment.

```10 people died in train accident```

In the first lecture, we saw that some people would code this as **negative**, where others consider it **neutral**. 

* The people that coded **negative** argued that dying is clearly a negative thing, thus implicitly defining sentiment as whether the news event itself is positive or negative. Let's call this the **event-based sentiment** camp.
* The people that coded **neutral** argued that as a news headline, it is a neutral description of the event. This implicitly defined sentiment as a subjective, emotional expression. Let's call this the **subjective sentiment** camp. 

Both interpretations are fair. 
As a researcher, it is your responsibility to define what you mean by sentiment in your study, and instruct the coders properly so that their coding is valid for answering your research question.
In this assignment, we'll practice this using the following three headlines:

```{r}
headline1 <- "10 people hurt in a plane crash."
headline2 <- "10 people hurt in a terrible plane crash."
headline3 <- "10 people hurt in a surprisingly harmless plane crash!"
```

Your goal is to create instructions to code these headlines for event-based and subjective sentiment. You can add any instruction you want (definition of sentiment, examples). The only exception is that you are not allowed to mention the specific event (10 people hurt in accident) in the instruction, because that would be cheating.

> **2.1**. Create an instruction for measuring **event-based sentiment**. Use this instruction with our `sentiment_classifier` to code the three headlines. Report your code and the results. 

> **2.2**. Reflect on how well your **event-based sentiment** instruction worked. Did you get the results your expected? Why do you think your instruction worked (or did not work)?  

> **2.3**. Create an instruction for measuring **subjective sentiment**. Use this instruction with our `sentiment_classifier` to code the three headlines. Report your code and the results. 

> **2.4**. Reflect on how well your **subjective sentiment** instruction worked. Did you get the results your expected? Why do you think your instruction worked (or did not work)?  



# 3. Coding Economic Sentiment Headlines

Now that we have a gold standard and a classifier, we can test how well our classifier performs! You already saw how to do this last week. Here we show it again with a few extra tricks. Make sure to follow along with the example, because you'll be doing this yourself in Assignment 3.

## Validating the sentiment classifier

In the previous practical you learned how to use `purrr` to map over multiple texts and apply the sentiment classifier to each.
Here we'll do the same.
But since this time we're going to map over more texts, we want to make sure that if one text fails (e.g. the Gemini API throws an error), that it doesn't break the whole process. 
Otherwise, you might have just coded 200 texts, but then if text 201 throws an error you would have to start all over again!

To prevent this, we'll add a safety net around `sentiment_classifier` that tells it to return `NA` (missing value) in case of an error.
This can easily be done with the `possibly` function from the `purrr` package.
Here we create a new function `safe_sentiment_classifier` that works just like `sentiment_classifier` if it succeeds, but returns `NA` in case of an error.

```{r}
library(purrr)
safe_sentiment_classifier <- possibly(sentiment_classifier, otherwise = NA)
```

And now we can use `map_vec` again.
To show a quick example, we'll take a random sample of 10 headlines from `gold` and code them.

```{r, eval=F}
gold_sample <- sample_n(gold, 10)

predicted <- map_vec(gold_sample$headline, 
                     safe_sentiment_classifier, 
                     instruction = "Code the economic sentiment of these news headlines",
                     .progress = TRUE)
```

And now we can compute the performance metrics using the `confusionMatrix` function, just like last time.
The only thing we need to do first is convert both our `predicted` values and the gold standard values (`gold$gold`) to `factors` with the same values.
For `predicted` we can simply use `as.factor` like last time.
But since the `gold` data is in a numeric format (`-1, 0, 1`) we'll use the `factor` function to explicitly map these values (`levels=`) to the right labels (`labels=`).

```{r, eval=F}
library(caret)

predicted = as.factor(predicted)
actual = factor(gold_sample$gold, 
                levels = c(-1, 0, 1), 
                labels = c('negative','neutral','positive'))

confusionMatrix(predicted, actual, mode='prec_recall')
```


## Assignment 3

Now we have everything in place to return to our initial goal: to see how well we can approach or even exceed the performance of the best machine learning models in [Van Atteveldt et al. (2021)](https://www.tandfonline.com/doi/pdf/10.1080/19312458.2020.1869198) using zero-shot classification. 


> **3.1**. Create the instruction that you'll use in the sentiment classifier. Include an explanation of why you think this is a good instruction. 

> **3.2**. Use the classifier to code all headlines in the gold standard (so not a sample!), and calculate the performance metrics. Write a detailed report of the performance. Report the accuracy of the model, and report the precision, recall and F1 scores per sentiment category. Also look at the confusion matrix to reflect on what types of categories were most often mixed up. Overall, how well do you think your model performed? 

*(Running the analysis in 3.2  can take a few minutes, so if you do not want to wait, you can already work on 3.3).*

> **3.3**. Summarize the performance metrics from table 2 in Van Atteveldt et al. (2021). What **manual content analysis approach** (manual/crowd coding) had the best accuracy, and what **automatic content analysis approach** (machine learning / dictionaries) had the best accuracy? Report the precision, recall and F1 scores for these two approaches.

> **3.4**. Compare these results to your own results. How well did your zero-shot classification analysis perform compared to the best automatic approach that Van Atteveldt et al. reported? How well does it perform compared to the best manual approach?
