---
title: 'Social Media Analytics V'
author: 'Sofia Gil-Clavel & Kasper Welbers'
subtitle: "Introduction to Large Language Models and Zero-Shot for Labelling"
output:
  html_document:
    theme: united
    highlight: tango
    css: custom.css
    toc: true
    toc_depth: 1
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: sentence
---

```{r opts, echo = FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.path = "img/",eval = FALSE)
```

VU Amsterdam

# 1. Preparing the R-script

It is recommended that you clean your R-environment before you start coding. So, let's do that.

1. Open a new R-script.
2. In the first lines, add the following code `rm(list=ls())` and `gc()`. 
3. Run the code

From the second step, `rm(list=ls())` cleans the R environment, so that you do not drag any variables from previous R-sessions. The second line (`gc()`) makes sure that those variables do not occupy your computers' RAM anymore.

**Exercise 1.a:** Clean the R-environment.

As you will be working with processes that can take some time to run, we recommend that you run those scripts and save the results either in your computer or in your JupyterHub folder as a csv or as a table. 

**Remember that if you need more resources, you can always use the course environment in [JupyterHub](https://hub.compute.vu.nl/).**

**Also, you can check this [blog post](https://societal-analytics.nl/blogs/20250201_computing-power/) to learn how to upload data to and download data from JupyterHub.**

# 2. Introduction to today's practical

In the previous practical, you used the set of packages offered by `tidyverse`. In specific, you used `dplyr` for data handling. `ggplot2` for data visualization. `stringr` for text handling. 

The main goal of this practical is for you to learn to use Large Language Models (LLMs) to annotate data. For this, you will use the package `tidyllm`, `purrr`, and `caret`. `purrr` is part of the `tidyverse` packages.

During practical 3, you used boolean queries, which is a rule-based technique for labeling texts. Back then you learned that a big limitation of rule-based approaches is that we just can’t seem to come up with a sufficiently complex set of rules to systematically annotate the text we want to analyse.

So, how can we use a computer to do something without specifically telling it how to do it? This is where LLMs come in.

For example, consider the case of a spam filter. We want to use a computer to automatically label new emails as spam or not spam. We first need to create our Gold Standard. For this we collect some old emails, and manually label them as either spam or not spam. So now we can give the computer a set of emails that are spam, and a set of emails that are not spam, according to our own interpretation. From this data, we can test and compare different LLMs to see which one performs the best in this task. Then, when it receives new emails, we can continue to use such LLM to decide whether an email is more likely to be spam or not spam.

The cool thing about LLMs, or GenAI, is that they may work for many different types of data. The input can be a set of emails that we want to classify into either spam or not spam, but it can also be a set of animal pictures that we want to classify into which animal is depicted. Generally speaking, GenAI lets you use computers to perform complicated tasks without having to tell them how to perform these tasks. The price is that you need to very carefully test the models and think about the inherent costs of using them. For example, did you properly test whether the model is the best for the task? Who would be accountable if the model performs poorly? Have you considered the risks of using a public model? What are the pros and cons of using the models locally in your computer or from a server, e.g. JupyterHub?

During this tutorial, we will give you the tools to use LLMs in the best possible way, while considering the inherent costs and risks of implementing them. For this, you will use the R-packages: 

* `tidyverse`
* `tidyllm`
* `caret`

**Hint:** Remember that if you do not know how to use a function, you can either type `? NAME_OF_THE_FUNCTION` or you go to the "Help" panel and search for the function. This will display all the information related with the function, as well as some examples on how to use it. You can also check the package's [Cheat Sheet](https://posit.co/resources/cheatsheets/).

**Exercise 2.1:** In your R-script download and load the packages. For this you need to use the function `install.packages()` and `library()`.



# 3. Accessing Large Language Models (LLMs)

In this section, you will interact with LLMs in two ways:

* A public big LLM provider: [Gemini](https://ai.google.dev/gemini-api/docs/api-key).
* An LLM on your computer: [Ollama](https://ollama.com/).

You will reflect on the pros and cons of each and you will also learn to put them in production and to test their results.

To access LLMs, we need to access their Application Programming Interfaces (APIs). In general, an API is used for computer programs to communicate with each other (for a more detailed explanation please check the tutorial [Use APIs to source your data in R](https://raw.githack.com/Delft-RCafe/resources/main/themes/apis_in_r/apis-in-r.html)). Yes, this means that, in general, any time you prompt an LLM, you are actually interacting with an API. 

## 3.1 A public big LLM provider

As you know there are several companies that provide different LLMs. For example, Gemini is developed by Google DeepMind; Chat-GPT is developed by OpenAI; Llama is developed by Meta Platforms, Inc; and so far, and so on. Each company has their own parameters to access their LLMs via thier APIs. So, you would need to look for each in order to access them. To avoid the hassle, we will use the package `tidyllm` which already has those parameters loaded.

In this tutorial, we will demonstrate how to access Gemini from Google DeepMind. This is because Google Gemini has a free tier. The steps to get your API key are simple:

* Go to [https://aistudio.google.com/app/api-keys](https://aistudio.google.com/app/api-keys).
* Sign in and click the "create api key" button at the top-right. Follow the steps for making a project and key. **You don't have to register any payment details.**
* Once you have created your key you should see a column "Billing tier", which should say "free tier".  
* Copy your API key.

The following chunk shows you the way you would access Google Gemini using `tidyllm` and its `gemini_chat()` function:

```{r Using tidyllm, eval=FALSE}
Sys.setenv(GOOGLE_API_KEY = "YOUR API KEYo")

llm_message("Hi, how are you?") |>
  gemini_chat()%>%
  get_reply()



```

You can also select a specific model to use. Models are different in terms of speed, costs and performance, so you often want to pick the best model for your job. When using LLMs for simple content analysis tasks of many documents, it is often good to pick a model that is relatively fast and cheap. You can find an overview of the available models [here](https://ai.google.dev/gemini-api/docs/models)

```{r}
llm_message("Hi, how are you?") |>
  gemini_chat(.model='gemini-3.1-flash-lite-preview')%>%
  get_reply()
```

The number of times you can use a model, and the number of 'tokens' you use, is limited depending on your subscription. Since we use the Free tier, we can only use every model for a little bit. At the time of writing, the limit for the free tier of the "gemini-3.1-flash-lite-preview" model is max 15 requests per minute, and 500 request per day. You can keep track of how much you used and how much is available [here](https://aistudio.google.com/rate-limit).

The reason we're showing you this is that it's important that you understand a bit about the current economy and infrastructure for using LLMs. Even though you can use websites like ChatGPT and Gemini for free, when you use LLMs at scale for analyzing data, you will typically pay on a usage basis, and need to respect the rate limits (i.e. how many requests you can make within a given period of time).

### Creating your own 'coder' function

In this tutorial you will learn how to create your own **coder**. Essentially, we're building a robot to do content analysis for us!

To make our robot easy to use, you'll learn how to create your own *functions*. You've seen many functions by now, and you've seen how you can add more functions by installing and opening packages. But one of the powerful features of R is that you can also make your own functions quite easily.

In its basic form, this is what a function looks like. We basically tell R: Create a function named `my_fun`. This function has two parameters: `parameter1` and `parameter2`. With `parameter2 = 'default value'` we specify a default value (i.e. the value if the users doesn't give a value). In the **function body**, the part between the { } brackets, we put the code that the function will perform using these parameters. And in the `return` function, we specify what the function should output.

```{r first function, eval=FALSE}
## Create a function with two parameters
my_fun <- function(parameter1, parameter2 = "default value"){
  ## Here you put the code that the function performs
  
  ## You return the results like this
  return("Variables to return")
}
```

Its easier to understand with an example. Here we create a function that takes two numbers as parameters, and returns their sum. Notice that within the function *body* (the part between the { } brackets) we can use `number_1` and `number_2`, which represent the input from the user. 

```{r example function, eval=FALSE}

add_numbers <- function(number_1, number_2 = 20){
  result = number_1 + number_2 
  return(result)
}

add_numbers(10, 40)

## If we don't specify the second argument, it uses the default (20)
add_numbers(10)


```

Now let's use this to create a function that takes a prompt and model as input, and returns the response from Gemini. 

```{r GOOGLE_GEMINI}
my_first_robot <- function(prompt, model = 'gemini-3.1-flash-lite-preview') {
  result = llm_message(prompt) |>
    gemini_chat(.model=model)%>%
    get_reply()
  
  return(result)
}

my_first_robot("Say hi, robot!!")

## optionally, specify a different model
my_first_robot("Say hi, robot!!", model = 'gemini-2.5-flash-lite')
```


**Exercise 3.a:** Test the function `my_first_robot()` with two different prompts and two different models. Report the code, prompts and responses. 


**Exercise 3.b:** In the second lecture we talked about the pros and cons of using professional LLM providers (from big tech companies). Reflect on the risks and limitations of using a professional LLM provider for content analysis.

**Exercise 3.c:** Describe a research scenario where it would be OK to use an LLM from a professional LLM provider.



## 3.2 An LLM on your computer

Sometimes, it is not possible (for legal, ethical or technical reasons) to use a professional LLM provider. Luckily, it is easy nowadays to run an LLM on your own computer!! For this tutorial we'll work with lightweight LLMs, so they're not as good as the big ones. But once you know how to do it, the only thing you need to perform a professional level analysis is a bigger computer.

To work with LLMs locally on our own computers, we will install "Ollama", which is a local AI model management software. Ollama grants you full control to download, update, and delete models easily on your system. This prioritize strict data security, as we do not pass data to companies like OpenAI. In addition to basic management, Ollama lets you track and control different model versions. So we can switch between different models. 

You can download Ollama here: [https://ollama.com/](https://ollama.com/). Click on the download button, choose the right operating system. Install the software after it has downloaded. Once you have installed it, open it, for the `tidyllm` package to call it, the software has to be running.

Next, we'll need to install a model to use. You can find a list of available models here: [https://ollama.com/search](https://ollama.com/search). Keep in mind that models require both RAM and hard drive. So, we suggest that, if your computer has 8Gb of RAM, then you should only work with the 1b to 1.5b version of the models. As they take about 1GB on your computer and should run with most hardware.

If you have enough space on your hardware and your computer has at least 12GB RAM, you can consider larger models, between 2b and 3b (~2GB) which should perform better. Only if you have a really strong computer, we would suggest to download something bigger, but you must still take the time to double check whether your computer has enough RAM and hard drive to allocate them. 

**IMPORTANT:** If you decide to use JupyterHub for this practical, then it is not necessary to neither download Ollama nor to download the models. They both are already loaded in the JupyterHub course environment. 

To download models, you can use the function `ollama_download_model()`. In this tutorial we'll use and compare 2 models: `llama3.2:1b` and `deepseek-r1:1.5b`. Go ahead and run the following lines of code. This might take a few minutes depending on your device and internet connection.

```{r ollama download model, eval=FALSE}
ollama_download_model("llama3.2:1b")
ollama_download_model("deepseek-r1:1.5b")
```

To interact with the software, we will continue to use `tidyllm`. Note that the code is pretty much identical to what we used above for Gemini. The only difference is that in the chat function we use the `ollama` function, where we also provide specific parameters for Ollama (like `.temperature`, which controls the randomness of the model.). 

**IMPORTANT:** In the JupyterHub course environment, for the `ollama()` function (inside `chat()`), you need to add the parameter `.ollama_server = "http://ollama.local:11434"`.


```{r ollama prompt}
llm_message("Hi, how are you?")%>%
  chat(ollama(.model = "llama3.2:1b",
              .temperature = .0))%>%
  get_reply()
```


**Exercise 3.2.a:** Write your own function (like `my_first_robot` above), and call it `ollama_local`. This time it has 3 parameters: in addition to the `prompt` and `model`, add a `temperature` parameter. Give the temperature a default value of 0. The function should return the reply.

**Exercise 3.2.b:** Test your function with two different prompts and the different two models: `llama3.2:1b` and `deepseek-r1:1.5b`. Keep the temperature set at 0.



## 3.3 Using `map_vec()` from the package `purrr` 

Now that you have a function to call the LLM, we can use it to send multiple requests. This is an important step towards building our coder robot. What we're aiming for, is to have a script that can send thousands of texts to our robot to code for us. 

For this, we will use the function `map_vec()` from the package `purrr`. `map_vec()` allows you to perform the same function multiple times given a vector (i.e. variable) of inputs. 

As an example, here we use it to send six different prompts sequentially to the `ollama_local` function that you created in 3.2.1. 

```{r map_vec, eval=FALSE}
# Create a vector with multiple prompts
prompts <- c(
  "say hi",
  "say cheese",
  "say no",
  "say yes",
  "say more",
  "say less"
)

replies <- map_vec(prompts,             # Pass the vector of prompts
                   ollama_local,        # Pass the name of the function
                   model="llama3.2:1b", # Pass the model paramater (of ollama_local)
                   temperature=0.5,     # Pass the temperature parameter
                  .progress = T)       # This adds a nice progress bar

# Now we can use tibble to create a dataframe/tibble from both vectors
tibble(prompts, replies)
```

Now that we can make multiple requests at once, let's also use this to explore what that **temperature** argument does. To investigate this, we're going to ask the model to randomly say either "positive" or "negative" multiple times.

```{r}
prompt <- 'Respond with the word "positive" or "negative". respond only with one word'
prompts <- rep(prompt, 10) ## repeat it 10 times

replies <- map_vec(prompts,             
                   ollama_local,        
                   model="llama3.2:1b", 
                   temperature=1,   
                  .progress = T)    

# Now we can use tibble to create a dataframe/tibble from both vectors
tibble(prompts, replies)
```


**Exercise 3.3.a:** Now, use the model `llama3.2:1b` and the same prompt, but set the temperature to 0. Save the results in a tibble. 

**Exercise 3.3.b:** What does the `temperature` parameter do? What is the difference between a low temperature (0) and a high temperature (1)? 


**Exercise 3.3.c:** Is the LLM always returning only the word "positive" or "negative"? Is the LLM always doing what you asked? 



# 4. Using LLMs to annotate data

Now, that you have a function to call different models and a way to pass different prompts, we are going to annotate some data. The data is the "sentiment_tweets.csv". You can find this data both in CANVAS (same data as practical 4) and the SMA JupyterHub environment (`/srv/scistor/data/SMA/data/machine_learning_tweets/sentiment_tweets.csv`)

**Hint:** Use the function `read.csv()`.

**Exercise 4.a:** Open the "sentiment_tweets.csv" data and save it in the variable TWEETS.

**Exercise 4.b:** Clean the text using the functions you learned in practical 4. Remember to the save the new data in itself, so that changes are saved.


This dataset contains around 60,000 tweets, but for this assignment we will only take a small sample of 30 tweets. Once you know how to perform the analysis with 30 tweets, you can also perform it for all tweets, but it would just take a lot more time.

To create the sample, you will use the function `sample_n()`. To make your code reproducible, you need to call the function `set.seed(123)` before calling the function `sample_n()`. In short, `set.seed(any number you want)` sets a starting point (a seed) for random functions. By setting the same seed before drawing the random sample, you will always get the same sample. 

```{r sample of 30, eval=FALSE}
## import and clean tweets
## TWEETS <- read.csv(...)
## ... (cleanup steps)

set.seed(123)
tweets_sample <- TWEETS %>%
  sample_n(size = 30)
```

## 4.1 Zero-Shot Classification

Now, we are going to use a technique called "Zero-Shot Classification". This means that the model will not need any training (i.e. it has seen zero training examples). We'll simply ask the model to 'code' a text and provide instructions, just like we would ask a person to do this in manual content analysis.

To create our zero-shot classifier (i.e. our imitation of a human coder) we'll take the following steps:

**First**, we create a general instruction that specifies the coding task. 

```{r the annotator instruction}
instruction = llm_message('You are an expert at manual content analysis. I will show you a tweet, and you will respond with the sentiment of that tweet, using the labels "negative", "neutral" and "positive"') 
```

**Second**, we specify a **structured output**. as you saw in section 3.3, sometimes LLMs are not very good at following instructions. We might ask it to respond with "negative", but it will sometimes say "Negative", "NEG", or "0". This is problematic, because for our analysis we need it to be consistent.

In order to force the model to respond with a specific output, we'll use a **structured output**. Simply put, we'll strictly tell the model how it can respond to our question. This is a powerful technique for using LLMs for content analysis, that is supported by most (but not all!) LLM providers.

A structured output looks as follows. We create a schema in which we specify that the model should return a 'sentiment' label, which should be a factor with the value 'negative', 'neutral' or 'positive'. For this assignment we are only interested in this sentiment, but to showcase how usefull structured outputs are, we'll also ask it to code the 'topic' of the tweet, without giving any instructions.

```{r the annotator schema}
structured_output = tidyllm_schema(
  sentiment = field_fct("The tweet sentiment", c("positive", "neutral", "negative")),
  topic = field_chr("The topic of the tweet"),
)
```

**Third**, we can make the request. Note that we first send the instruction, and then send the `llm_message` with the tweet. We pass the structured_output to the .json_schema parameter. 

Also note that this time we use `get_reply_data()` instead of `get_reply()`, because this parses the result from the structured output for us. The `result` in this code is a named vector with the variables that we specified in the structured output schema (sentiment, topic). We can use `result$sentiment` to specifically get only the sentiment code

```{r}
result <- instruction |>
  llm_message("This is an example of a very happy tweet")%>%
  chat(ollama(.model = "llama3.2:1b",
              .json_schema = structured_output))%>%
  get_reply_data()


result
result$sentiment
```

Your exercise is now to put all these parts together in a custom zero-shot classification function. This function just needs to have 1 parameter: `tweet`. All the other components (selecting the model, creating the instruction, specifying the structured output) can happen inside the function body (the part between the { } brackets).

**Exercise 4.1.a:** Create a function (like you did in 3.3) that takes a `tweet` as input, and returns the sentiment as output. Use the `llama3.2:1b` model. You can give the function any name you like (e.g. `sentiment_classifier`). So you should be able to call `sentiment_classifier("Bob hates skeeing")`, and this should return the sentiment ("negative").

**Exercise 4.1.b**: Use your function to classify the first three tweets in the TWEETS data frame.


## 4.2 Labeling the sample using the different models

Now we have all the building blocks for our coder robot:

* you have a function (4.1.a) to label the sentiment of a text
* you know how to use the function `map_vec()`

Now you'll use `map_vec()` with this function, similar to 3.3. Instead of sending the prompts, you now send the tweets (`tweets_sample$text`) as the first argument 

**Exercise 4.2.a:** Label the 30 tweets in the `tweets_sample` data using map_vec. From the result, create a tibble that has the columns `tweet` and `sentiment`. 

**Exercise 4.2.b:** Now add the `tweets_sample$label` column to the tibble, and name it `real_sentiment`. If you did everything correctly, you should now have a tibble with 3 columns: the `tweet`, the `real_sentiment` from the data, and the `sentiment` as coded by your zero-shot classifier. report this tibble. 


# 5. Evaluating the model

The final step is to validate how well our zero-shot classifier performed. For this we'll again use the precision, recall and F1 measures that we learned in practical 3. But this time, we'll let R calculate them for us instead of calculating them by hand.

Here is an example for how to use the `confusionMatrix` function from the `caret` package. This function takes two factor vectors as input that represent the actual values (i.e. the gold standard) and the predicted values. 


```{r factors}
library(caret)

example <- tibble(
  classifier = c('positive','positive','negative','neutral','positive'),
  gold_standard = c('positive','neutral','negative','neutral','negative')
)

actual = as.factor(example$gold_standard)
predicted = as.factor(example$classifier)
confusionMatrix(predicted, actual, mode = 'prec_recall')
```


**Exercise 5:** Use the confusionMatrix function to calculate the precision, recall and F1 values. Then reflect on how well our zero-shot classifier performed, and report the performance metrics in your own words (so don't just copy/paste the result). Be detailed. Also use the confusion matrix to reflect on what types of mistakes were most common.

Note that since we used a small sample size and didn't really fine-tune the prompt, the performance is pretty random, and you might happen to have quite low scores. But that's ok for todays assignment. Now that you've learned how to build and test a classifier, it is relatively simple to increase the sample size and experiment with different prompts. We'll turn to that challenge in the next practical.
