---
title: 'SMA Project'
subtitle: "Additional instructions and R code"
output:
  html_document:
    theme: united
    highlight: tango
    css: custom.css
    toc: true
    toc_depth: 2
editor_options:
  chunk_output_type: console
---

```{r opts, echo = FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
```

VU Amsterdam

# Week 2 - Manual Content Analysis

In week 2 of the project in P6 you will perform a manual content analysis in order to create a gold standard.
To understand the goal, please have another look at the [Google Sheet](https://docs.google.com/spreadsheets/d/18oquhpDp3uyRzP0jcLbkBXOUKoLcCtRo8m73AodKnFM/edit?usp=sharing) that we used in Practical 6, which contained the gold standard in the [Van Atteveldt et al. (2021)](https://www.tandfonline.com/doi/pdf/10.1080/19312458.2020.1869198) paper.

This time, you will create this Google Sheet yourself (and we strongly recommend also using Google Sheets for this).
Notice that there are multiple tabs in the Van Atteveldt data: **combined**, **coder_1**, **coder_2**, etc.
Each sheet has the same texts, in the same order.
In **coder_1-3**, each respective coder independently coded the text according to the codebook. 
The reason for doing this in separate sheets is that they do not see the codes of the other coders, because their judgments need to be independent of each other.



## Creating your own Google Sheet

For the project you will create your own Google Sheet. 
You can use the Van Atteveldt sheet as an example.

For example, here is [another Google Sheet](https://docs.google.com/spreadsheets/d/1aRToQRcCJXMTEhiReEWpS8622hedj3U5GWi6mD6gh3A/edit?usp=sharing) where we used the same format, but this time with (fake) **Tweets about MacDonalds**. For every tweet, three coders (anna, bob and charles) each coded two measures: **topic** (food taste, customer service, price) and **sentiment** (-1, 0, 1).

To create your own sheet:

* Make a copy of the example
* Copy/paste the data you gathered into it. Make sure to give every text a unique ID. 
* Create columns for your measures
* Create different tabs for each team member


## BONUS: Analyzing images

You are also allowed to use images as units of analysis. 
In this case, you should not directly copy/paste the images into Google Sheets.
This would work for the manual content analysis, but it's hard to read, and makes things more complicated for the zero-shot analysis (as shown below).

Instead, you can [use this format](https://docs.google.com/spreadsheets/d/1uThubEUmFGTcVQ4dyC4JksJWt-pB2a_QKcNR4m5A6sY/edit?usp=sharing)

Instead of texts, you'll put the file names in the Google Sheet.
Then you'll also create a folder (like Google Drive) with the actual images, [like this](https://drive.google.com/drive/folders/1VbOzNI1nPlfRx7qNAxcjE2z364_q_i4D?usp=sharing)

It's important that the names in this column are **identical** to the real file names, because this allows us to automate the zero-shot classifier,
as explained in the mini tutorial (to be published soon) on how to do zero-shot classification of images.

## Performing the content analysis

Once everything is in place, you can perform the content analysis. 
Every coder studies the codebook, and tries coding the texts.

Do not immediately begin coding everything.
Make sure to try some texts together to see if the codebook is clear and you generally agree.
Ideally, also perform a pilot test, where you first code a small number of texts and then calculate intercoder reliability to see to what extent you agreed.
If the intercoder reliability of the pilot is low, you'll first need to update the codebook.


## Intercoder reliability

Here's a small recap of intercoder reliability in R.
We'll use the example data with multiple measures (topic and sentiment).
If you want to follow along with the examples, make sure to download the Google Sheet **combined** tab as a CSV file first (just like in Practical 6)


```{r}
library(tidyverse)

## Make sure to change the file location if it's different for you
intercoder <- read_csv("~/Downloads/manual_content_analysis_example - combined.csv")
```

### Interval variables

In practical 6 you already saw how we can calculate Krippendorf's alpha for **interval** variables.

```{r}
library(irr)

intercoder |>
  select(anna_sentiment:charles_sentiment) |>
  t() |>
  kripp.alpha(method = 'interval')
```

### Nominal variables

The **topics** in our example data are **nominal**, so there we'd use `method = 'nominal'`. 
However, the `kripp.alpha` function prefers if values are numeric, so we first create `id` columns.
That is, we'll say that Food Taste has ID 1, Customer service has ID 2, etc.
We'll do this by first making the variable a `factor` with the correct `topic_levels`,
and then making them numeric.
It looks a bit verbose, but if you compare `anna_topic` with `anna_topic_id` I'm sure you'll get the gist of it.

```{r}

## Specify topic levels. The order determines the IDs (Food Taste = 1, etc.)
topic_levels = c('Food Taste', 'Customer Service', 'Price')

## Conver the topic character vector into a numeric ID vector
intercoder <- intercoder |>
  mutate(
    anna_topic_id = as.numeric(factor(anna_topic, levels = topic_levels)),
    bob_topic_id = as.numeric(factor(bob_topic, levels = topic_levels)),
    charles_topic_id = as.numeric(factor(charles_topic, levels = topic_levels))
  )

intercoder |>
  select(anna_topic, anna_topic_id) |>
  head()
```

Now we can use these IDs in the `kripp.alpha` function.

```{r}
intercoder |>
  select(anna_topic_id:charles_topic_id) |>
  t() |>
  kripp.alpha(method = 'nominal')
```

Since you'll be working with 4 or 5 coders, it can also be useful to inspect alpha scores with different selections of coders. If you have a low score among all coders, it might just be one coder that is very idiosyncratic. You could then consider dropping this coder from the gold standard creation task, and transparently explaining this in the method section (because it's not ideal to drop a coder, and in a real study, it would be better to figure out why this coder was so different and try to resolve the issue.)

```{r}

## Anna VS Bob
intercoder |>
  select(anna_topic_id, bob_topic_id) |>
  t() |>
  kripp.alpha(method = 'nominal')

## Anna VS Charles
intercoder |>
  select(anna_topic_id, charles_topic_id) |>
  t() |>
  kripp.alpha(method = 'nominal')

## Bob VS Charles
intercoder |>
  select(bob_topic_id, charles_topic_id) |>
  t() |>
  kripp.alpha(method = 'nominal')
```
  
  
## Creating the gold standard

Now for the final step, you only need to create the **gold** variable(s) in the Google Sheet. This is just like the **gold** column in [The Google Sheet from Practical 6](https://docs.google.com/spreadsheets/d/18oquhpDp3uyRzP0jcLbkBXOUKoLcCtRo8m73AodKnFM/edit?usp=sharing). Except maybe you'll have multiple gold columns if you had more measures (e.g. gold_topic, gold_sentiment).

You need to describe how you determined the gold value in cases of disagreement (for an example, see how Van Atteveldt et al say they did it). You could for instance use majority vote, or discuss cases of disagreement until you reach an agreement. 

# Week 3 - Automatic Content Analysis

Here we recap the code from practical 6. 
We'll also do things slightly differently, so that you can also choost to have the classifier code multiple variables at once. 

## Set up TidyLLM

```{r, eval=F}
library(tidyllm)
Sys.setenv(GOOGLE_API_KEY = "[ENTER API KEY HERE]")


llm_message("Motivate me for working hard on my assignment. Tell me how proud you would be if I ace it.") |>
  gemini_chat(.model = "gemini-2.5-flash-lite") |>
  get_reply() |>
  cat()
```

```{r, echo=F}
## use this to make the knit work. Just make sure to remove the key afterwards
library(tidyllm)
Sys.setenv(GOOGLE_API_KEY = "")
```

## Creating a classifier

Just like in practical 6, you'll need to create a classifier function.
This time we'll make it a bit more flexible.
We'll make it so that we can also change the output_schema, and so that it can code multiple variables at once by returning a tibble with a column for every variable.

Another small difference is how we create the prompt. This time we use `sprintf`, which takes a string and replaces `%s` with the next arguments. Here we use it to put the instruction and text together in a format that LLMs understand well (i.e. markdown).


```{r}
classifier <- function(text, instruction, schema) {
  prompt <- sprintf("## Instruction\n\n%s\n\n## Text\n\n%s", 
                    instruction, text)
  
  llm_message(prompt) |>
    gemini_chat(.model = "gemini-2.5-flash-lite", 
                .json_schema = schema,
                .temperature = 0) |>
    get_reply_data() |>
    as_tibble() |>
    mutate(input_text = text)
}
```

Now we can create any instruction and define any schema for what the classifier should code.

```{r}
instruction <- "You are an expert coder in a manual content analysis. The goal is to classify Tweets about MacDonalds into different consumer satisfaction topics, and code the sentiment towards this topic"

schema = tidyllm_schema(
  topic = field_fct("The consumer satisfaction topic", 
                    c("Food Taste", "Customer Service", "Price")),
  sentiment = field_fct("The sentiment towards the topic", 
                    c("positive", "neutral", "negative"))
)

classifier("Dirty tables everywhere", instruction, schema)
```

## Coding multiple texts

In practical 6 we showed you how to map over multiple values.
Here we'll do the same thing, but instead of returning NA on an error, it returns a tibble. This way it works well with the new tibble output of the classifier.

```{r}
safe_classifier <- possibly(classifier, otherwise = tibble(.error=T))
```

And when mapping over values, we before used `map_vec`, because we had just one variable (i.e. one vector). This time the results are tibbles, so we need to bind them all together as rows, which we do by first using `map` and then `bind_rows` over the result.

```{r, echo=F}
set.seed(1)
```

```{r, cache=TRUE}
## here we just copy the intercoder data to gold, to clarify its the gold standard
gold <- intercoder

rows <- map(gold$Tweet,
    safe_classifier,
    instruction = instruction,
    schema = schema,
    .progress = T)

result <- bind_rows(rows)
result
```

## Calculating precision and recall

And given the results, this step is basically the same.
The only difference is that we need to select which variable to use.

```{r}
library(caret)

sentiment_levels <- c('negative','neutral','positive')

predicted =  factor(result$sentiment, levels = sentiment_levels)
actual = factor(gold$gold_sentiment, 
                levels = c(-1, 0, 1), 
                labels = sentiment_levels)

confusionMatrix(predicted, actual, mode='prec_recall')
```


## BONUS: Analyzing images

If you decide to analyze images as units of analysis, the process is very similar. 
The only difference is that instead of giving TidyLLM a text, we'll give it the location of an image on our computer. 
The only thing that makes this a bit more complicated than using texts is that you need to be more comfortable with finding files on your computer.

**!! IMPORTANT**: Make sure to have the latest version of `tidyllm` installed for this feature.

For this example we'll use the example of [cat images](https://docs.google.com/spreadsheets/d/1uThubEUmFGTcVQ4dyC4JksJWt-pB2a_QKcNR4m5A6sY/edit?usp=sharing).
To try it yourself, you also need to download the [folder of cat images](https://drive.google.com/drive/folders/1VbOzNI1nPlfRx7qNAxcjE2z364_q_i4D?usp=sharing)

```{r}
images <- read_csv("~/Downloads/using_images_example - combined.csv")
images
```

```{r}
image_classifier <- function(image_file, instruction, schema) {
  llm_message(instruction, .media = img(image_file)) |>
    gemini_chat(.model = "gemini-2.5-flash-lite", 
                .json_schema = schema,
                .temperature = 0) |>
    get_reply_data() |>
    as_tibble() |>
    mutate(image_file = image_file)
}
```

We use it in the same way. We just provide the file path instead of a text.

```{r}
instruction <- "You are an expert coder and a cat whisperer. Your job is to tell the sentiment that a cat is expressing. Code the sentiment of the cat in this image"

schema = tidyllm_schema(
  sentiment = field_fct("The sentiment of the cat", 
                    c("positive", "neutral", "negative"))
)

image_classifier("~/Downloads/cats/angry_cat.jpeg", instruction, schema)
```

Now, the trickiest part (but not thaaat tricky). 
We need to loop over the file names, and then get the file from the correct directory.
In my case, I downloaded the files to the directory `~/Downloads/cats`, so I'll need to add this directory to the file names.

```{r}
full_path <- file.path("~/Downloads/cats", images$image)
full_path
```

And then we use this in the `map` function.

```{r}
safe_image_classifier <- possibly(image_classifier, otherwise = tibble(.error=T))

rows <- map(full_path,
    safe_image_classifier,
    instruction = instruction,
    schema = schema,
    .progress = T)

result <- bind_rows(rows)
result
```

And that's how easy it is to perform automatic image analysis these days :).
